---
title: "chapter_4_probabilistic_classifiers_with_bayes"
author: "jason grahn"
date: "2023-11-21"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr)
```

"Bayes" = principles for describing the probability of events and how those probabilities are revised with additional information.

Refresher: 

* probability measured as a number between 0 and 1 (0 to 100%)
* low = not likely to occur, high = likely to occur.

Bayes Classifiers use training data to find probability of outcomes based on featurd-based evidence.
Classifier then applied to unlabeled data to predict most likley class from the new example. 

Use these when info from numorous attributes should be considered at the same time, to extimate overall probability of an outcome. 

Most ML Algos ignore weak effect features, Bayes uses them all and adjusted subtly. Lots of small features could have a combined impact! 

Bayes in a nutshell: likelihood of an event (outcome) should be based on evidence across many opportunities for the events to occur.

Examples: 
* heads / tails
* rainy weather
* winning the lottery
* spam text

**Probability**

$\frac{successes}{attempts}$

denoted as 

The Probability that "A" event occurred: $P(A)$ 
Probability that rain occured? $P(rain)$

The inverse is probably that rain did not occur: $1 - P(rain)$ aka the complement $P(A^c)$ (Probability of the compliment of event A)

**Joint Probability**

How the probabilty of one event is related to the probability of the other. (and we use the upside-down "U" symbol to denote these intersections.)

When "Spam" and "Viagra" occur, then the _intersection_ between $P(spam) \cap P(viagra)$ is a joint probability. 

Quote from the text: **dependent events** are the basis of predictive modeling. Just as the presence of clouds is predictive of a rainy day, the appearance of the word Viagra is predictive of a spam email.

$P(A \cap B) = P(A)*P(B)$

# Computing conditional probability with Bayes’ theorem

$P(A|B) = \frac{P(A \cap B)}{P(B)} $

The probability of _Event A_ happened given _Event B_ also happened.

Bayes says: The best estimate of $P(A|B)$ is the proportion of trials where A occurred with B out of _all_ events where B occurred. 

which we can flip to $P(A|B) = \frac{P(A \cap B)}{P(B)} = \frac{P(B|A)*P(A)}{P(B)}$ using the transitive property of multiplication! 

"Prior probability" is the known. In the case of the spam filter and viagra supplements, its easier to measure the $P(spam)$. 

Then we can do _posterior probability_ to measure how likely a message is to be spam. Greater than 50%? more likely spam! 

$P(spam|viagra) = \frac{P(viagra|spam) * P(spam)} {P(viagra)}$ 

P(spam|viagra) = posterior probability
P(viagra|spam) = likelihood
P(spam) = prior probability
P(viagra) = marginal likelihood 

So... $posterior = \frac{likelihood * prior}{marginal}$

# the Naive Bayes algo

strengths
* simple, fast, effective
* works well with noise, missing data, & lots of features
* few examples needed to train
* easy to get estimated probability

weaknesses
* often-faulty assumptions of equal importance and independence (rarely true)
* not great with numeric features
* estimates not as reliable as predicated classes

## Classification with Naive Bayes

**class-conditional independence** = events are independent so long as they are conditioned on the same class value. Allows us to multiply individual conditional probabilities instead of looking at joint probability.

"The probability of spam is equal to the likelihood that the message is spam divided by the likelihood that the message is _either_ spam or ham."
`r round(0.012 / (0.012 + 0.002),3)`

The probability of ham is equal to the likelihood that the message is ham divided by the likelihood that the message is eitehr. 
`r round(0.002 / (0.012 + 0.002),3)`

Build a frequency table. 
build a likelihood table.
multiply the conditional probabilities with the naive assumption of independence.
divide by the total likelihood -- transforms each class likelihood into a probability.

## The LaPlace estimator

when we've never seen something before, it's prior is zero. So when we do a 'normal' Bayesian classifier multiplying the conditional probabilities, zero times zero = zero. The **Laplace estimator** adds a small number to each of the counts, which ensures that each feature has a non-zero probability of occuring. Typically it's set to 1 so all classes-features have at least _1_. (later notes: 1 is almost always used.)

You add the laplace estimator in both the numerator and denominator of each likelihood.

## Using numeric features

Because naive bayes users frequency tables, each feature _must_ be categorical. Numeric features don't have categories! 

* make the numeric features discrete; that is, put them into bins (aka: **Binning**). 
++ Bin these according to logical conclusions about the data. 
++ a clock has 24 discrete hours, or maybe 4 bins of time per day "early morning" "late morning" "afternoon" evening". 
++ use a histogram! 
* if you can't figure our how to bin, just pop 'em into quantiles. 

# example, filtering spam texts

## 1, collecting data

```{r}
sms_raw <- read_csv("https://raw.githubusercontent.com/PacktPublishing/Machine-Learning-with-R-Fourth-Edition/main/Chapter%2004/sms_spam.csv")

# the file is also in the /data folder
```
## 2 exploration & prep

```{r}
str(sms_raw)
dplyr::glimpse(sms_raw)
```

```{r}
sms_raw$type <- factor(sms_raw$type)

str(sms_raw)

table(sms_raw$type)
```

We use the `tm` package to text mining.
```{r}
library(tm)
```

We're going to create a corpus. A corpus is a collection of text documents. The text "documents" can be of any length. Use `VCorpus()` for this, the "v" meaning "volatile" (we'd used `PCorpus()` to access a permanent corpus on a database). 

```{r}
sms_corpus <- VCorpus(VectorSource(sms_raw$text))

sms_corpus
```

A corpus is a list, so we can use list tools to pick documents out of it. 

```{r}
inspect(sms_corpus[1:2])
```
And to see what's _in_ those.. 

```{r}
as.character(sms_corpus[[1]])
```

ok so let's look at multiple documents...

```{r}
lapply(sms_corpus[1:2], as.character)
```

Lots of problems with text strings... Punctuation and character cases are two main ones. So let's try to clean this up a bit. 

```{r}
sms_corpus_clean <- 
    tm_map(sms_corpus,
           content_transformer(tolower))

# did it work? 
as.character(sms_corpus[[1]]) == as.character(sms_corpus_clean[[1]])
as.character(sms_corpus_clean[[1]])
```

Notice that everything is lower case now. Next we remove numbers (and these iterative steps are things that seem better wrapped in `mutate()`)

```{r}
sms_corpus_clean <- tm_map(sms_corpus_clean, removeNumbers)

lapply(sms_corpus[1:5], as.character)
lapply(sms_corpus_clean[1:5], as.character)
```

Then we remove stop words, and may as well include puntuation next..

```{r}
sms_corpus_clean <-
    tm_map(sms_corpus_clean,
           removeWords, 
           stopwords())

sms_corpus_clean <-
    tm_map(sms_corpus_clean,
           removePunctuation)
```

... Note that removing punctuation can change a `hello...world` into `helloworld`

We need to install the `SnowballC` package to do stemming sooo that's being done on the side.. Stemming strips `ed`, `ing`, and plurals from words so we can count the root words.

```{r}
sms_corpus_clean <-
    tm_map(sms_corpus_clean, stemDocument)

# and lets strip whitespaces too..

sms_corpus_clean <- 
    tm_map(sms_corpus_clean, stripWhitespace)
```

What does this end up looking like? 

```{r}
lapply(sms_corpus[1:5], as.character)
lapply(sms_corpus_clean[1:5], as.character)
```

### data prep - text docs to words

**tokenization**: a single element of a text string, words. 

so we split the 'document' (text strings) into their individual words, then count how many times those words occur in each document. This produces with a massively wide data, as we attempt to count _each_ word of the entire corpus for each document. We end with a **sparce matrix** where most of the counts are zero. 

We use the default settings of `DocumentTermMatrix()`.

```{r}
sms_dtm <- DocumentTermMatrix(sms_corpus_clean)
```

At least some of that prep we did earlier could've been done here too. When we do it this way, we notice the mix of cases in the `tm` package. (there's probably a better way out there to do this that has more attention to detail...).

```{r}
sms_dtm2 <- 
    DocumentTermMatrix(
        sms_corpus,
        control = list(
            tolower = TRUE,
            removeNumbers = TRUE,
            stopwords = TRUE,
            removePunctuation = TRUE,
            stemming = TRUE))
```

Though, doing it this way provides slightly different results. 

```{r}
sms_dtm
sms_dtm2
```

```{r}
# lets get rid of this because we aren't going to use it.
rm(sms_dtm2)
```

(It would be nice if the results output was a table so we could compare the output values against each other.)

## 3. training & test sets

we'll split 75% training, 25% testing. 
data is already randomized, so we're just going to take a lop off the top for training.

```{r}
sms_dtm_train <- sms_dtm[1:4169,]
sms_dtm_test <- sms_dtm[4170:5559,]

# and save a vector of their labels. 
# this is not how i like to do this kind of thing, because it relies on manual input, 
# which is prone to mistypes and errors! 

sms_train_labels <- sms_raw[1:4169,]$type
sms_test_labels <- sms_raw[4170:5559,]$type
```


```{r}
prop.table(table(sms_train_labels))

# tibble(sms_train_labels) %>%
#     count(sms_train_labels) %>% 
#     mutate(freq = n / sum(n))
```

```{r}
prop.table(table(sms_test_labels))
```

## 4. Visualizing with word clouds

and of course we need a new `wordcloud` package. i'm installing that through the console. 

The idea here is that different words should **POP** between the spam & ham data.

```{r}
library(wordcloud)
```

```{r}
wordcloud(sms_corpus_clean, 
          min.freq = 50, 
          random.order = FALSE)
```

Since this shows _all_ the messages, we need to force a subset then cloud again (again... this method doesn't scale!)

```{r}
spam <- subset(sms_raw, type == "spam")
ham <- subset(sms_raw, type == "ham")
```

```{r}
wordcloud(spam$text, max.words = 40, scale = c(3, 0.5))
wordcloud(ham$text, max.words = 40, scale = c(3, 0.5))
```

## 5. creating indicator features for frequent words

we need to reduce the number of "features" (columns) because that sparse matrix sucks. we'll get ride of any word that appears in less than 5 messages. We save that as it's own vector.

```{r}
sms_freq_words <- findFreqTerms(sms_dtm_train, 5)

str(sms_freq_words) 

## 1137 different terms make the cut!
```

Let's apply this vector as a filter to our training data! 

```{r}
sms_dtm_freq_train <- sms_dtm_train[ ,sms_freq_words]
sms_dtm_freq_test <- sms_dtm_test[ ,sms_freq_words]
```

_now_ we convert the numeric features (the word counts) into categorical features (`Yes` or `No` strings, which would probably be better as TRUE/FALSE strings instead). Over in Tidyland, we'd probably get away with a mutate_all that does this through a case statement? 

```{r}
convert_counts <- function(x) {
    x <- ifelse(x > 0, "Yes", "No")
}
```

```{r}
sms_train <- apply(sms_dtm_freq_train,
                   MARGIN = 2, # applies the function to the columns, (MARGIN = 1 applies the function to rows)
                   convert_counts)

sms_test <- apply(sms_dtm_freq_test,
                   MARGIN = 2,
                   convert_counts)
```

## 6. training the model

we use the `naivebayes` package which I also installed through the console. 

```{r}
library(naivebayes)
```

```{r warning=FALSE}
sms_classifier <- naive_bayes(sms_train, sms_train_labels)

# muting errors because the book says I dont need to worry about this for now. 
# the error come from using the defalt leplace estimator
```

```{r}
sms_test_pred <- predict(sms_classifier, sms_test)
```

```{r}
gmodels::CrossTable(sms_test_pred, 
                    sms_test_labels,
                    prop.chisq = FALSE,
                    prop.c = FALSE,
                    prop.r = FALSE,
                    dnn = c('predicted', 'actual'))
```

## 7. improving the model

```{r}
sms_classifier2 <- naive_bayes(sms_train,
                               sms_train_labels,
                               laplace = 1)

# see, no errors when we make laplace = 1
```

```{r}
sms_test_pred2 <- predict(sms_classifier2, sms_test)
```

```{r}
gmodels::CrossTable(sms_test_pred2, 
                    sms_test_labels,
                    prop.chisq = FALSE,
                    prop.c = FALSE,
                    prop.r = FALSE,
                    dnn = c('predicted', 'actual'))
```

Notice that this reduced the number of false positives (ham messages classified as spam) as well as false negatives (spam messages classfied as ham). This model is 3 pp better than 0. 

